He-Man: A domain specific language for lightweight concurrency

### 2012-03-11 paper/paper.tex
Work on related work section.

To test the suitability of our system for more complicated programs
and to demonstrate separating network logic and application logic, we
built a very simple HTTP server. The network logic is written in
He-Man and request parsing and response header building is written in
C. The C request parsing is implemented as a function that takes a
buffer and returns whether it forms a complete HTTP request and, if
so, fills a buffer with the file name to serve. The entire server is
68 lines of He-Man Haskell and 39 lines of C. It suffers from a number
of major deficiencies, but is complete enough to benchmark. To make
sure that it does not get away with sending less data than the servers
it is being compared against, it sends some ``padding'' headers with
useless data.

Our benchmark measurements are taken using weighttp, an http
benchmarking tool associated with lighttpd
\cite{weighttp}. Measurements were collected by requesting a file many
times with a variable number of concurrent connections; the experiment
was repeated with a 100 byte file (80000 requests), a 16 kilobyte file
(30000 requests), and a 128 kilobyte file (8000 requests).
weighttp was run with 4 threads making requests.
Since the test repeatedly requests the same file, it should always be in
the kernel's disk cache.

For our experiments, both the client and server ran Linux kernel
2.6.32 on quad core 3.1 GHz Intel Core i5-2400 processors with 8
gigabytes of RAM. The machines were connected over a 100 megabit
Ethernet network.
We ran the tests against our He-Man HTTP server, Apache 2.2.14 (installed
with the default Ubuntu 10.04 LTS settings, except for client, thread,
and process limits being adjusted upwards) and Lighttpd 1.4.26
(installed with the default Ubuntu 10.04 LTS settings).

Each trial was run 10 times; the plots show the arithmetic mean of the
number of requests per second the servers handled; error bars show the
standard deviation.

Figure \ref{fig:graph100} compares the three servers while serving a
very small (100 byte) file. Measurements with small files test the
overhead of the web server more clearly than measurements with large
files, which wind up testing how efficiently the kernel can transfer
data. In this test, He-Man outperforms Apache and lighttpd across all
of our measurements. This can be attributed to factors both in He-Man
and in our server. He-Man imposes very low overhead and very low
connection creation cost, which makes it easy to quickly handle many
new connections. Our server is also much less featureful that Apache
and lighttpd, which means it needs to do less. It is difficult to
separate out these factors.

Figure \ref{fig:graph128} and Figure \ref{fig:graph16} compares our
server to Apache and lighttpd for 128 KB and 16 KB files. On these
loads, our server performs nearly identically to Apache while being
outperformed by the highly-turned lighttpd. In these tests, both
Apache and He-Man begin to handle fewer requests per second above some
number of concurrent requests (about 400 for the 16 KB file and 300
for the 128 KB file). We investigated the He-Man httpd server's
behavior in these tests, and found that the measured slowdown in
requests/second occured when, after rapidly processing most of the
requests in a test batch, a few requests would spend a very long time
waiting to connect, causing a pause at the end of the run. During
these pauses, both the client and the server were blocked in the
kernel, waiting for network events. This strongly suggests to us that
the problem is not with the He-Man compilation phase but that our web
server would benefit from more careful tuning of network interactions
(which may require improvements to the runtime).

\begin{figure*}[!ht]
\centering
\includegraphics[width=0.9\textwidth]{graph_100.pdf}
\caption{Web server benchmark, 100 byte file.}
\label{fig:graph100}
\end{figure*}
\begin{figure*}[!ht]
\centering
\includegraphics[width=0.9\textwidth]{graph_16K.pdf}
\caption{Web server benchmark, 16 KB file.}
\label{fig:graph16}
\end{figure*}
\begin{figure*}[!ht]
\centering
\includegraphics[width=0.9\textwidth]{graph_128K.pdf}
\caption{Web server benchmark, 128 KB file.}
\label{fig:graph128}
\end{figure*}

@@

% Related work:
%\cite{VanderwaartCrary} re bounded time
%\cite{vonBehrenEtAl} re Capriccio
%\cite{vonBehrenConditBrewer} re cooperative multithreading
%\cite{LiZdancewic}
%\cite{LauerNeedham} also intro
%\cite{MarlowEtAl} also intro

% Functional programming:
%\cite{Vinoski}
%\cite{Hellstrom}
%\cite{BiagioniEtAl}
%\cite{LiEtAl}
%\cite{GraunkeEtAl}
%\cite{Queinnec}

@@

@@

A number of functional languages support lightweight threads with
user-space scheduling, most notably Erlang, which was designed for
highly concurrent telecommunications applications, and easily scales
to 20,000 clients \cite{Hellstrom}. The Haskell runtime supports a
variety of concurrency primitives which permit a lightweight threaded
programming model \cite{LiEtAl}.

@@

% XXX taken from old intro

Languages like Erlang and Haskell provide lightweight threads via user-space
schedulers. This affords even better performance, but remains unnecessarily
powerful for this problem space, imposing runtime overhead on applications which
don't require preemptable, independently scheduled threads \cite{Vinoski}.

>>>>

To test the suitability of our system for more complicated programs
and to demonstrate separating network logic and application logic, we
built a simple (and bad) HTTP 1.0 server. The network logic is written
in He-Man and request parsing is written in C. The C side is
implemented as a function that takes a buffer and returns whether it
forms a complete HTTP request and, if so, fills a buffer with the
file name to serve. The entire server is 54 lines of He-Man Haskell and
27 lines of C. It suffers from a number of major deficiencies, but is
complete enough to benchmark.

Our benchmark measurements are taken using ab, the Apache HTTP server
benchmarking tool \cite{ApacheAB}. Measurements were collected by
having a variable number of client threads request a file 100 times;
the experiment was repeated with a 16 kilobyte file and a 128 kilobyte
file. We ran the tests against both our He-Man HTTP server and Apache
2.2.20, installed with the default Ubuntu 11.10 settings, except for
client, thread, and process limits being adjusted upwards. Since the
He-Man runtime is currently only single-threaded, we ran the Apache
web server pinned to a single processor. Since the test repeatedly
requests the same file, it will always be in the kernel's disk cache,
and thus this test is not disk bound.

For our experiments, the server ran Linux kernel 3.0.0 on a Lenovo
Thinkpad W520 with a quad core 2.3 GHz Intel Core i7-2820QM and 8
gigabytes of RAM, and the client ran on a Lenovo Thinkpad T500 with a
Intel Core 2 Duo and 8 gigabytes of RAM.  % XXX: client specs
The machines were connected by a gigabit Ethernet link.
Unfortunately, ab is not particularly scalable, so we had trouble
testing with more than 500 clients.

Figure \ref{fig:graph128} compares our server to Apache for a 128 KB
file. Our server performs nearly identically to Apache on this load.
Figure \ref{fig:graph16} compares our server to Apache for a 16 KB
file. In this scenario, we substantially outperform Apache. We
conjecture this is because with a smaller file size, overhead for
connection setup begins to dominate, and connection setup in He-Man
consists of a few memory allocations whereas connection setup under
Apache may require spawning new OS level threads.

\begin{figure*}[!ht]
\centering
\includegraphics[width=0.9\textwidth]{graph_128K.pdf}
\caption{Web server benchmark, 128 KB file.}
\label{fig:graph128}
\end{figure*}
\begin{figure*}[!ht]
\centering
\includegraphics[width=0.9\textwidth]{graph_16K.pdf}
\caption{Web server benchmark, 16 KB file.}
\label{fig:graph16}
\end{figure*}
\begin{figure*}[!ht]
\centering
\includegraphics[width=0.9\textwidth]{graph_100.pdf}
\caption{Web server benchmark, 100 B file.}
\label{fig:graph100}
\end{figure*}

@@

% Related work:
%\cite{VanderwaartCrary} re bounded time
%\cite{vonBehrenConditBrewer} re cooperative multithreading
%\cite{LiZdancewic}

@@

\subsection{Hybrid concurrency models}

The Lauer-Needham duality \cite{LauerNeedham} between threaded and event-driven
concurrency has given rise to a number of programming models attempting to
bridge the gap between these abstractions.

Capriccio \cite{vonBehrenEtAl} is a user-level threading library which
implements cooperative (non-preemptive) multithreading, and replaces blocking
I/O primitives in \t{libc} with a wrapper for non-blocking I/O. Capriccio allows
programmers to write C code in a threaded style, and performs a compiler
analysis to statically allocate per-thread stack space. Capriccio's goals are
similar to ours, but we feel that He-Man's approach of compiling to source is
more portable and much simpler.

Flux \cite{BurnsEtAl} is a domain-specific language intended, like He-Man, to
handle concurrency in network applications separately from the rest of the
application's logic. Flux allows programmers to declaratively specify how data
flows through the server's components, and compiles to a threaded or
event-driven runtime which implements this behavior concurrently. Flux source
code cannot express loops, except for an implicit outermost loop, as a way of
preventing deadlocks; He-Man allows more expressive control flow while making
the same guarantee.

Many functional programming languages present a native lightweight thread
abstraction. Erlang \cite{VirdingEtAl} was designed for highly concurrent
telecommunications applications, and easily scales to 20,000 clients
\cite{Hellstrom}. Concurrent ML \cite{ReppyEtAl} adds concurrency primitives to
Standard ML, implemented with front-end continuations, and a back-end
synchronization protocol. The GHC Haskell runtime supports concurrency
primitives \cite{LiEtAl} by scheduling lightweight threads onto a small number
of kernel threads \cite{MarlowEtAl}.

% XXX: remains unnecessarily general for us?

While He-Man's goals are similar to
those of several other systems, we feel that it solves

\subsection{Lightweight threads}

% Functional programming:
\cite{Vinoski}
\cite{Hellstrom}
FoxNet for Standard ML \cite{BiagioniEtAl}
\cite{GraunkeEtAl}
\cite{Queinnec}

@@

@@
